Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2015-02-02 draft_lm1.tex
Update table an figure 3 ( closes #15).
also improved plotting code.
applied UK spell checker

In this paper, we propose a new approach to estimate the number (fraction) of 
susceptibles using a simplified model of dengue transmission based on a 
single-strain Susceptible-Infectious-Removed (SIR) model with time-varying 
infection rate.
In order to bypass the limitations of not knowing the serotype-specific 
seroprevalence and the exact behavior of the force of infection through time, we
propose to inform the time-varying transmissibility using the \rr series 
derived from the notification data~\cite{nishiura}.
We extend a Bayesian framework previously used to estimate the number 
of susceptibles in Influenza epidemics in Europe~\cite{pone2011} to include 
time-varying force of infection and derive a probability distribution  for 
\rr to accommodate uncertainty in the estimates. 
Then, from the incidence series and the population at risk, we  calculate the 
attack ratio for each epidemic.
We apply our method to estimate $S_0$ before every major dengue epidemic in the 
city of Rio de Janeiro, Brazil in the last 18 years.

@@

\subsection*{Mathematical modeling} % The model

@@

For our analysis we chose to let the force of infection vary with time, just 
as it does in the actual epidemics, as seen in the data. So as the epidemic 
progresses, the effective transmission  rate changes and is 
given by 
\begin{equation} 
 \label{eq:effbeta}
 \beta(t) = \frac{R_t\cdot\tau}{S_0}
\end{equation}
where $R_t$ is the effective reproductive number, estimated as 
in~\ref{eq:Rtestimate}.
The complete model with the time-verying force of infection is given by
the system of ordinary differential equations:
\begin{align}
   \label{eq:model}
 \frac{dS}{dt} &= -\beta(t)SI \\     \nonumber
 \frac{dI}{dt} &= \beta(t)SI - \tau I&\\      \nonumber
 \frac{dR}{dt} &= \tau I&
\end{align}  
where $S + I + R = 1 \: \forall\: t$. % \in [T_0, T_1]$. 
Of course, this is a rather simplified model, in which, for instance, the 
vector is omitted.
The rationale for this simplification is based on the ability of the 
empirically derived $R_t$  to incorporate the effects of the fluctuating vector 
populations.
Also, although there are multiple circulating serotypes, our approach
can not discriminate between them due to the lack of serotype-specific data.
Nevertheless, this modeling strategy can still provide some insight into the 
disease dynamics and allows us to estimate the initial fraction of susceptibles 
$S_0$, a key epidemiological parameter.

@@

Methods for estimating the  number of susceptibles have been proposed 
before, for other diseases~\cite{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}.
These methods try to reconstruct the entire 
series of infectious and susceptibles for measles 
outbreaks from case data.
In the case of dengue, the full (multi-year/multi-epidemic)
series of susceptibles to all possible serotypes, cannot be reconstructed based 
solely on a deterministic transmission model, since the arrival/re-emergenge of 
new serotypes (which are a stochastic events) can change drastically the pool 
of susceptibles throwing off any sequential estimation based on the incidence
dynamics.

@@

Figure \ref{Fig:S0}, shows the model from (\ref{eq:model}) fitted to the data. 
In it we can see that the susceptibles series in each epidemic starts at the 
estimated level of $S_0$.
The proportion of susceptibles may seem low, but we 
must remember that these estimates are being affected by an unknown 
underreporting factor, which experts suggest is somewhere between 5 and 10, 
i.e. for every case observed there are 5 or 10 unobserved.
Since this underreporting affects both the numerator and denominator of 
(\ref{eq:AR2}), its effects should cancel out, giving us an unbiased attack 
ratio 
estimate.
One other possible source of bias which would lead to the 
underestimation of $S_0$ could come from a significant part of the population 
not being exposed to the disease.
However, as we can see in figure \ref{fig:mapas} , the disease spreads pretty 
much homogeneously over 
the entire city (in the last four epidemics, at least).

@@

The authors hope that this evidence will motivate public health 
authorities to invest in annual serological surveys, to determine the 
susceptibility profile to each dengue virus as well as to estimate the 
underreporting factor of the notification system.

\section*{Acknowledgments}
LMC is grateful to Dr. Leonardo Bastos for useful discussions on the 
posterior inference for \rr.
The authors are also grateful to Claudia T. 
Code\c{c}o for helpful discussions about the manuscript.

\section*{Appendix}
\subsection*{A remark on prior distributions and tail behavior of the 
distribution of $R_t$}
\label{sec:tails}
There are a number of approaches to deriving the distribution of \rr.
Alternatively to the approach described in the main text~\cite{mantel}, one 
could use the conditional distribution of \rr on 
$Y_{t+1}$ and $Y_t$ as defined in equation A7 of Nishiura et al.\cite{nishiura}:
\begin{equation}
\label{seq:unorm}
f_{R}(R_{t}) = (Y_tR_{t})^{Y_{t+1}} e^{-Y_tR_{t}}
\end{equation}
Noticing the kernel of (\ref{seq:unorm}) is that of a gamma distribution with 
$a_2 = Y_{t+1}+1$ and $b_2 = Y_t$, we obtain a proper density from which to 
construct $c_{\alpha}(R_t)$, simply by computing the appropriate quantiles of 
said distribution.
 This density is
\begin{equation}
\label{seq:densityNishiura}
f_N(R_t| a_2, b_2) =  \frac{b_2^{a_2}}{\Gamma(a_2)} R_t^{a_2-1} e^{-b_2 R_t}
\end{equation}

In order to decide which approach to take, it may be of use analyzing the 
tail behavior of the derived distributions for \rr. 
Consider the case of using a flat $Uniform(0, 1)$ prior for $\theta_t$.
With $a_0 = b_0 = 1$, $a_1 = a_2$ and $b_1 = b_2 + 1$.
The beta prime (inverse beta distribution) will have heavier tails compared to 
the conditional distribution proposed by~\cite{nishiura}, thus providing more 
conservative confidence/credibility intervals.
To see that one needs simply take the ratio of the Beta prime and Gamma 
(unnormalized) densities and evaluate the limit as $R_t$ goes to infinity:
\begin{equation}
 \label{seq:densityratio}
 \lim_{R_t\to\infty}\frac{f_P(R_t| a_1, b_1)}{f_N(R_t| a_2, b_2)} =  
\lim_{R_t\to\infty}\frac{e^{Y_{t}R_t}}{(1 +R_t)^{Y_{t} + Y_{t +1}+2}} = \infty
\end{equation}
Finally, note that we deliberately construct $c_{\alpha}(R_{t})$ as a 
equal-tailed $100\alpha\%$ credible set, rather than a less conservative 
highest posterior density (HPD) interval.

@@

An advantage of the Bayesian approach is that one can devise prior 
distributions for $\theta_t$ taking advantage of the intuitive parameterization 
and flexibility of the beta family of distributions.
Prior elicitation can also be done for \rr and the hyper-parameters directly 
plugged into the prior for $\theta_t$. 
One can, for example, choose prior mean and variance for \rr and find $a_0$ 
and $b_0$ that satisfy those conditions.
Let $m_0$ and $v_0$ be the prior expectation and variance for $R_t$. 
After some tedious algebra one finds
\begin{align}
\label{seq:elicitation}
a_0 &= \frac{m_0v_0 + m_0^3 + m_0^2}{v_0} \\
b_0 &= \frac{2v_0 + m_0^2 + m_0}{v_0}
\end{align}
If one wants only to specify $m_0$ and the coefficient of variation $c = 
\sqrt{v_0}/ m_0$ for $R_t$ \textit{a priori}, some less boring 
algebra gives:
\begin{align}
\label{seq:elicitationcv}
a_0 &= \frac{m_0^3c^2 + m_0^3 + m_0^2}{m_0^2c^2} \\
b_0 &= \frac{2m_0^2c^2 + m^2 + m}{m_0^2c^2}
\end{align}

@@

\begin{table}[!ht]
\caption{
{\bf Median attack ratio and 95\% credibility intervals calculated according to 
(\ref{eq:AR2})}. Values are presented as percentage of total population. 
$^\dag$: Year corresponds to the start of the epidemic. 
Sometimes the peak of cases occur in the following year.
$^\ddag$: Susceptible fraction.}
\begin{center}
\begin{tabular}{c|c|c}
\hline
Year$^\dag$ & Median attack ratio (\%) & $S_0^\ddag$ (\%)\\
\hline
1996 & 38.64 (17.18-54.19) & 0.171\\
1997 & 87.31 (73.83-87.31) & 0.273\\
1998 & 49.86 (49.19-49.86) & 0.142\\
1999 & 10.76 (3.72-20.24) & 0.345\\
2000 & 25.49 (24.47-26.98) & 1.547\\
2001 & 48.12 (46.57-49.15) & 4.951\\
2005 & 14.74 (10.23-21.38) & 1.472\\
2006 & 11.22 (8.00-14.41) & 2.810\\
2007 & 15.08 (14.76-15.38) & 13.451\\
2008 & 13.73 (3.10-30.84) & 0.672\\
2010 & 18.30 (17.34-19.33) & 4.540\\
2011 & 8.59 (8.20-9.36) & 21.482\\
2012 & 13.92 (12.71-14.92) & 6.208\\

>>>>

In this paper, we propose a new approach to estimate the number (fraction) of 
susceptibles using a simplified model of dengue transmission based on a 
single-strain Susceptible-Infectious-Removed (SIR) model with time-varying 
infection rate.
In order to bypass the limitations of not knowing the serotype-specific 
seroprevalence and the exact behaviour of the force of infection through time, 
we
propose to inform the time-varying transmissibility using the \rr series 
derived from the notification data~\cite{nishiura}.
We extend a Bayesian framework previously used to estimate the number 
of susceptibles in Influenza epidemics in Europe~\cite{pone2011} to include 
time-varying force of infection and derive a probability distribution  for 
\rr to accommodate uncertainty in the estimates. 
Then, from the incidence series and the population at risk, we  calculate the 
attack ratio for each epidemic.
We apply our method to estimate $S_0$ before every major dengue epidemic in the 
city of Rio de Janeiro, Brazil in the last 18 years.

@@

\subsection*{Mathematical modelling} % The model

@@

For our analysis we chose to let the force of infection vary with time, just 
as it does in the actual epidemics, as seen in the data. So as the epidemic 
progresses, the effective transmission  rate changes and is 
given by 
\begin{equation} 
 \label{eq:effbeta}
 \beta(t) = \frac{R_t\cdot\tau}{S_0}
\end{equation}
where $R_t$ is the effective reproductive number, estimated as 
in~\ref{eq:Rtestimate}.
The complete model with the time-varying force of infection is given by
the system of ordinary differential equations:
\begin{align}
   \label{eq:model}
 \frac{dS}{dt} &= -\beta(t)SI \\     \nonumber
 \frac{dI}{dt} &= \beta(t)SI - \tau I&\\      \nonumber
 \frac{dR}{dt} &= \tau I&
\end{align}  
where $S + I + R = 1 \: \forall\: t$. % \in [T_0, T_1]$. 
Of course, this is a rather simplified model, in which, for instance, the 
vector is omitted.
The rationale for this simplification is based on the ability of the 
empirically derived $R_t$  to incorporate the effects of the fluctuating vector 
populations.
Also, although there are multiple circulating serotypes, our approach
can not discriminate between them due to the lack of serotype-specific data.
Nevertheless, this modelling strategy can still provide some insight into the 
disease dynamics and allows us to estimate the initial fraction of susceptibles 
$S_0$, a key epidemiological parameter.

@@

Methods for estimating the  number of susceptibles have been proposed 
before, for other diseases~\cite{bjornstad_dynamics_2002, 
wallinga_reconstruction_2003}.
These methods try to reconstruct the entire 
series of infectious and susceptibles for measles 
outbreaks from case data.
In the case of dengue, the full (multi-year/multi-epidemic)
series of susceptibles to all possible serotypes, cannot be reconstructed based 
solely on a deterministic transmission model, since the arrival/re-emergence of 
new serotypes (which are a stochastic events) can change drastically the pool 
of susceptibles throwing off any sequential estimation based on the incidence
dynamics.

@@

Figure \ref{Fig:S0}, shows the model from (\ref{eq:model}) fitted to the data. 
In it we can see that the susceptibles series in each epidemic starts at the 
estimated level of $S_0$.
The proportion of susceptibles may seem low, but we 
must remember that these estimates are being affected by an unknown 
under-reporting factor, which experts suggest is somewhere between 5 and 10, 
i.e. for every case observed there are 5 or 10 unobserved.
Since this under-reporting affects both the numerator and denominator of 
(\ref{eq:AR2}), its effects should cancel out, giving us an unbiased attack 
ratio 
estimate.
One other possible source of bias which would lead to the 
underestimation of $S_0$ could come from a significant part of the population 
not being exposed to the disease.
However, as we can see in figure \ref{fig:mapas} , the disease spreads pretty 
much homogeneously over 
the entire city (in the last four epidemics, at least).

@@

The authors hope that this evidence will motivate public health 
authorities to invest in annual serological surveys, to determine the 
susceptibility profile to each dengue virus as well as to estimate the 
under-reporting factor of the notification system.

\section*{Acknowledgements}
LMC is grateful to Dr. Leonardo Bastos for useful discussions on the 
posterior inference for \rr.
The authors are also grateful to Claudia T. 
Code\c{c}o for helpful discussions about the manuscript.

\section*{Appendix}
\subsection*{A remark on prior distributions and tail behaviour of the 
distribution of $R_t$}
\label{sec:tails}
There are a number of approaches to deriving the distribution of \rr.
Alternatively to the approach described in the main text~\cite{mantel}, one 
could use the conditional distribution of \rr on 
$Y_{t+1}$ and $Y_t$ as defined in equation A7 of Nishiura et al.\cite{nishiura}:
\begin{equation}
\label{seq:unorm}
f_{R}(R_{t}) = (Y_tR_{t})^{Y_{t+1}} e^{-Y_tR_{t}}
\end{equation}
Noticing the kernel of (\ref{seq:unorm}) is that of a gamma distribution with 
$a_2 = Y_{t+1}+1$ and $b_2 = Y_t$, we obtain a proper density from which to 
construct $c_{\alpha}(R_t)$, simply by computing the appropriate quantiles of 
said distribution.
 This density is
\begin{equation}
\label{seq:densityNishiura}
f_N(R_t| a_2, b_2) =  \frac{b_2^{a_2}}{\Gamma(a_2)} R_t^{a_2-1} e^{-b_2 R_t}
\end{equation}

In order to decide which approach to take, it may be of use analysing the 
tail behaviour of the derived distributions for \rr. 
Consider the case of using a flat $Uniform(0, 1)$ prior for $\theta_t$.
With $a_0 = b_0 = 1$, $a_1 = a_2$ and $b_1 = b_2 + 1$.
The beta prime (inverse beta distribution) will have heavier tails compared to 
the conditional distribution proposed by~\cite{nishiura}, thus providing more 
conservative confidence/credibility intervals.
To see that one needs simply take the ratio of the Beta prime and Gamma 
(unnormalized) densities and evaluate the limit as $R_t$ goes to infinity:
\begin{equation}
 \label{seq:densityratio}
 \lim_{R_t\to\infty}\frac{f_P(R_t| a_1, b_1)}{f_N(R_t| a_2, b_2)} =  
\lim_{R_t\to\infty}\frac{e^{Y_{t}R_t}}{(1 +R_t)^{Y_{t} + Y_{t +1}+2}} = \infty
\end{equation}
Finally, note that we deliberately construct $c_{\alpha}(R_{t})$ as a 
equal-tailed $100\alpha\%$ credible set, rather than a less conservative 
highest posterior density (HPD) interval.

@@

An advantage of the Bayesian approach is that one can devise prior 
distributions for $\theta_t$ taking advantage of the intuitive parametrization 
and flexibility of the beta family of distributions.
Prior elicitation can also be done for \rr and the hyper-parameters directly 
plugged into the prior for $\theta_t$. 
One can, for example, choose prior mean and variance for \rr and find $a_0$ 
and $b_0$ that satisfy those conditions.
Let $m_0$ and $v_0$ be the prior expectation and variance for $R_t$. 
After some tedious algebra one finds
\begin{align}
\label{seq:elicitation}
a_0 &= \frac{m_0v_0 + m_0^3 + m_0^2}{v_0} \\
b_0 &= \frac{2v_0 + m_0^2 + m_0}{v_0}
\end{align}
If one wants only to specify $m_0$ and the coefficient of variation $c = 
\sqrt{v_0}/ m_0$ for $R_t$ \textit{a priori}, some less boring 
algebra gives:
\begin{align}
\label{seq:elicitationcv}
a_0 &= \frac{m_0^3c^2 + m_0^3 + m_0^2}{m_0^2c^2} \\
b_0 &= \frac{2m_0^2c^2 + m^2 + m}{m_0^2c^2}
\end{align}

@@

\begin{table}[!ht]
\caption{
{\bf Median attack ratio and 95\% credibility intervals calculated according to 
(\ref{eq:AR2})}. Values are presented as percentage of total population. 
$^\dag$: Year corresponds to the start of the epidemic. 
Sometimes the peak of cases occur in the following year.
$^\ddag$: Susceptible fraction.}
\begin{center}
\begin{tabular}{c|c|c}
\hline
Year$^\dag$ & median Attack Ratio & $S_0^\ddag$ \\
\hline
1996 & 0.39 (0.17-0.54) & 0.00171(0.0012-0.0038)\\
1997 & 0.87 (0.74-0.87) & 0.00273(0.0027-0.0032)\\
1998 & 0.5 (0.49-0.5) & 0.00142(0.0014-0.0014)\\
1999 & 0.11 (0.037-0.2) & 0.00345(0.0018-0.01)\\
2000 & 0.25 (0.24-0.27) & 0.0155(0.015-0.016)\\
2001 & 0.48 (0.47-0.49) & 0.0495(0.048-0.051)\\
2005 & 0.15 (0.1-0.21) & 0.0147(0.01-0.021)\\
2006 & 0.11 (0.08-0.14) & 0.0281(0.022-0.037)\\
2007 & 0.15 (0.15-0.15) & 0.135(0.13-0.14)\\
2008 & 0.14 (0.031-0.31) & 0.00672(0.003-0.024)\\
2010 & 0.18 (0.17-0.19) & 0.0454(0.043-0.048)\\
2011 & 0.086 (0.082-0.094) & 0.215(0.2-0.23)\\
2012 & 0.14 (0.13-0.15) & 0.0621(0.058-0.068)\\
